---
title: "S1 Code for: *Demography and climate influence sex-specific survival costs of reproduction over 60 years in a free ranging primate population*"
author: "Eve B Cooper, Lauren JN Brent, Noah Snyder-Mackler & James P Higham"
date: "`r Sys.Date()`"
output: 
  pdf_document:
    toc: true
    toc_depth: 3
    highlight: tango
---

# 1. Data loading and preparation

```{r setup, include=FALSE}
knitr::opts_chunk$set(tidy.opts = list(width.cutoff = 60), tidy = TRUE)
```

Loading Packages

```{r packages, include=TRUE,message=FALSE}
library(brms) #for all your modelling needs
library(tidyverse) #for data tidying
library(ggplot2) #for basic plotting needs
library(bayesplot) #for plotting Bayesian models easily in ggplot
library(tidybayes) #even tidier plotting
library(ggpubr) #format plots into figs for publication
library(readxl) #sometimes you have to use excel
```

Operating system and package versions

```{r sessionInfo}
sessionInfo()
```

Load all the data we will need:

```{r data load,message=FALSE}
#Rio Peidras weather station data
weatherRP <- read_excel("weather_rio_peidras.xlsx")

#import Humacao weather station data
weatherHum <- read_excel("weatherHum.xlsx")

#import census data files:
#dates associated with every birth and mating season on Cayo Santiago since 1959
yearSznSum<-read_excel("yearSznSum.xlsx")  
#monthly counts of population density
monthlyCensusDensity<-read_excel("monthlyCensusDensity.xlsx") 
#sex ratio during each calendar month of the study
monthlyCensusSexRatio<-read_excel("monthlyCensusSexRatio.xlsx")  

#import monthly survival records for all individuals included in the study:
survivalData<-read.csv("JAE_survival_data.csv")

#data cleaning:
#scale survival data (neccessary for some models, see below)
survivalDataScale<-survivalData %>%
  mutate(adultsTotal=c(scale(adultsTotal)))%>%
  mutate(Jdate=c(scale(Jdate)))%>%
  mutate(age=c(scale(age)))%>%
  mutate(birthSznSD=c(scale(birthSznSD)))%>%
  mutate(propFemale=c(scale(propFemale)))%>%
  mutate(tmax=c(scale(tmax)))%>%
  mutate(rain=c(scale(rain)))
```

# 2. Demographic and climatic variables over study period

In order to get a visual assessment of the variation in demographic variables (population density and sex ratio) and climatic variables (temperature and rainfall) over the course of the study period we will graph all these variables across time. We will also graph covariance between population density and sex ratio, and temperature and rainfall, in order to access their covariance.

This code creates Figure 1 in the manuscript:

```{r figure 1: variation across years, warning=FALSE,message=FALSE}
#demographic variables####
#plot population density
Gdens<-ggplot(filter(monthlyCensusDensity,adultsTotal>5),
              aes(x=year,y=adultsTotal))+ 
  geom_point(position="jitter")+
  labs(x="Year",y="Population Size", title="(A) Density")+
  theme_classic()

#plot sex ratio - females per male
monthlyCensusSexRatio<-mutate(monthlyCensusSexRatio, ratioFemale=females/males)
Gsex<-ggplot(filter(monthlyCensusSexRatio,adultsTotal>5),
             aes(x=year,y=ratioFemale))+
  geom_hline(yintercept=1,color="red")+
  geom_point(position="jitter")+
  labs(x="Year",y="Females per Male", title="(B) Sex ratio")+
  theme_classic()

#population density vs. sex ratio
monthlyCensusSum<-left_join(monthlyCensusDensity,
                            select(monthlyCensusSexRatio,month,ratioFemale),
                            by="month")
Gcomp<-ggplot(filter(monthlyCensusSum,adultsTotal>5),
              aes(x=adultsTotal,y=ratioFemale))+
  geom_point()+
  labs(x="Population Size",y="Females per Male",title="(C) Density/Ratio")+
  theme_classic()

#climatic variables####
#weather variables each year
#sort data legnthwise
birthData<-yearSznSum %>%
  select(Bszn, temp=meanTempBirth, Prcp =meanPrcpBirth)%>%
  mutate(season="birth")
mateData<-yearSznSum %>%
    select(Bszn, temp=meanTempMate, Prcp =meanPrcpmate)%>%
    mutate(season="mate")
fig1Data<-rbind(birthData,mateData)

#temperature
Gtemp<-ggplot(data=fig1Data,aes(x=Bszn,y=temp,group=season),fill=season)+
  geom_point(aes(color=season))+
  geom_smooth(aes(color=season,fill=season),method="lm")+
  labs(x="Year",y="Temperature (°C)", title="(D) Temperature")+
  scale_y_continuous(labels=function(x)x/10)+
  scale_fill_manual(values=c("red3","goldenrod"),name = "Season", 
                    labels = c( "Birth","Mating"))+
  scale_color_manual(values=c( "red3","goldenrod"),name = "Season", 
                     labels = c("Birth","Mating"))+
  theme_classic()


#rainfall
Gpre<-ggplot(data=fig1Data,aes(x=Bszn,y=Prcp))+
  geom_point(aes(color=season))+
  geom_smooth(aes(color=season,fill=season),method="lm")+
  labs(x="Year",y="Rainfall (mm)", title="(E) Rainfall")+
  scale_y_continuous(labels=function(x)x/10)+
  coord_cartesian(ylim=c(0,125))+
  scale_fill_manual(values=c("red3","goldenrod"),name = "Season", 
                    labels = c( "Birth","Mating"))+
  scale_color_manual(values=c( "red3","goldenrod"),name = "Season", 
                     labels = c("Birth","Mating"))+
  theme_classic()

#temperature by rainfall
weatherMonthly<-weatherRP%>%
  mutate(month=format(date, "%Y/%m"))%>%
  group_by(month)%>%
  filter(!is.na(tmax)| !is.na(prcp))%>%
  summarise(tmax=mean(tmax/10,na.rm=TRUE),prcp=mean(prcp/10,na.rm=TRUE),n=n())

Gweather<-ggplot(data=weatherMonthly,aes(x=tmax,y=prcp))+
  geom_point()+
  #geom_smooth(method="lm")+
  labs(x="Temperature (°C)",y="Rainfall (mm)", title="(F) Temp/Rainfall")+
  coord_cartesian(ylim=c(0,20))+
  theme_classic()

#dispaly figure 1
ggarrange(Gdens,Gsex, Gcomp,Gtemp,Gpre,Gweather)
```

# 3. Survival temporal trends

We ran 3 nested models for each sex. The most complex model for each sex (Model A) included a non-parametric effect of 'year' and an interaction between 'year' and 'reproductive investment' ('effort season':high [yes] vs. low [no]). For each sex we compared Model A to simpler models that were identical except for the exclusion of the interaction between 'year' and 'reproductive investment' (Model B), and the exclusion of both the interaction term and the non-parametric effect of 'year' (Model C).

## 3.1 Model formulations and results

The variables included in each model are:

Response variable:

-   **survival:** A binomial variable denoting whether the individual survived (1) or died (0) during the 30 day period.

Fixed effects:

-   **JDate:** Number of days from January 1st until the first day of the month survival is measured in. This is included as both first and second order terms using the 'poly()' function.

-   **BirthSznSD:** The total length in days of the mating or birth season currently being experienced, or experienced most recently

-   **effortSeason:** A 2-level categorical variable indicating if survival was measured during a month that majority overlapped with the individual's season of high reproductive investment ('yes') or not ('no').

-   **year:** a non-parametric spline term (only included in models A and B) denoting the year of study (1963 - 2022). Using a penalized thin-plate regression spline, this term estimates the relationship between year and the response variable monthly survival with a number of penalized additive smoothing functions determined by restricted maximum likelihood. For Model A, two splines are estimated, one for the season of high reproductive investment, and one for the season of low reproductive investment.

Random effect:

-   **individual_code:** The unique identifier for the individual who's monthly survival is being measured.

Below see the summary statistics and formulations for each model:

```{r survival temporal trends}
#we load previously run models since they each take several hours to run
#female models
survFemaleYear3<-readRDS("survFemaleYear3.rds") #model C
survFemaleYear2<-readRDS("survFemaleYear2.rds") #model B
survFemaleYear1<-readRDS("survFemaleYear1.rds") #model A
#male models
survMaleYear3<-readRDS("survMaleYear3.rds") #model C
survMaleYear2<-readRDS("survMaleYear2.rds") #model B
survMaleYear1<-readRDS("survMaleYear1.rds") #model A

#the priors for all models was set as follows:
#for intercept & population-level effects:
#weakly informative prior normally distributed w/ mean = 0, sd=1
#random effect prior had standard deviation follow exp. distribution w/ a rate of 1
prior = c(prior(normal(0, 1), class = "b"),
  prior(normal(0, 1), class = "Intercept"), 
 prior(exponential(1), class = "sd")) 

#temporal analysis model summary statistics########################################

#Female Temporal Survival Model A
survFemaleYear1$formula #model formula
summary(survFemaleYear1) #summary output

#Female Temporal Survival Model B
survFemaleYear2$formula #model formula
summary(survFemaleYear2) #summary output

#Female Temporal Survival Model C 
survFemaleYear3$formula #model formula
summary(survFemaleYear3) #summary output

#Male Temporal Survival Model A
survMaleYear1$formula #model formula
summary(survMaleYear1) #summary output

#Male Temporal Survival Model B
survMaleYear2$formula #model formula
summary(survMaleYear2) #summary output

#Male Temporal Survival Model C 
survMaleYear3$formula #model formula
summary(survMaleYear3) #summary output
```

## 3.2 Model diagnostic checks - example

Here we illustrate one example of model checking and diagnostics using the female temporal survival Model A. We ran similar model checks for all 10 models included in this study.


```{r Diagnostics Example}
#assessing model convergence
mcmc_plot(survFemaleYear1,type="dens_overlay")+
  theme(text = element_text(size = 5))
mcmc_plot(survFemaleYear1,type="trace")+
  theme(text = element_text(size = 5))
#chains have converged well
#estimates are normally distributed
#small variance of individual code random effect is expected for a survival model


#check for rhats deviating 10% larger or smaller than 1.00
mcmc_plot(survFemaleYear1,type="rhat_hist", binwidth=0.0001)
#no rhats larger than ~1.005 or below ~0.998
#indicates good convergence between chains


#autocorrelation
mcmc_plot(survFemaleYear1,type="acf_bar")+
  theme(text = element_text(size = 5))
#low autocorrelation at lag > ~5


#ratio of effective sample size to N
mcmc_plot(survFemaleYear1,type="neff_hist",binwidth=0.01)
#ratios of ESS to N range from 0.7 - 1.1, indicating satisfactory sampling


#posterior predictive check 
#simulates data from the posterior predictive distribution and compares to model predictions
pp_check(survFemaleYear1, ndraws = 500,type="bars")
#yrep is very similar to y, indicating robust predictions
```

## 3.3 WAIC model comparison

For each sex seperately, we compared Models A, B, and C using the widely applicable information criterion (WAIC), calculated using waic.brmsfit from the 'loo' package. When compared across a set of nested models run on the same data, a lower WAIC is indicative of a better fitting model based on the likelihood function. We then calculate the expected log pointwise predictive density (ELPD) between the model with the lowest WAIC and each subsequent model in order to quantify the difference in fit between models.


```{r WAIC model comparison}
#calculate model WAICs for later comparison using the 'waic' function from package 'loo'
waicF1<-waic(survFemaleYear1)
waicF2<-waic(survFemaleYear2)
waicF3<-waic(survFemaleYear3) 
waicM1<-waic(survMaleYear1)  
waicM2<-waic(survMaleYear2)  
waicM3<-waic(survMaleYear3)

#female Model comparison##################################################
waicF3 #female model C
waicF2 #female model B
waicF1 #female model A
loo_compare(waicF3,waicF2,waicF1)
#Model A has the lowest WAIC and significantly higher ELPD compared to models B and C
#this indicates that there is a significant non-parametric effect of year on female survival,
#and that this effect of year varies between the birth and non-birth seasons




#male Model comparison######################################################
waicM3 #male model C
waicM2 #male model B
waicM1 #male model A
loo_compare(waicM3,waicM2,waicM1)
#similar to in the females, Model A has the lowest WAIC,
#and significantly higher ELPD compared to models B and C
#this indicates that there is a significant non-parametric effect of year on male survival,
#and that this effect of year varies between the mating and non-mating seasons
```

## 3.4 Temporal trends best fit model graphs

Below we show the code to create Figure 2 in the manuscript. This figure illustrates the results of Model A (the best fit model), for each sex. It shows the non-parametric temporal survival patterns for females (Fig. 2A) and males (Fig. 2B) during seasons of high and low reproductive investment.


```{r Figure 2 prep,message=FALSE,warning=FALSE,fig.show='hide'}

#females - Model A plot####################################################

#create dataframe that is complete for model
femYearData<-survivalDataScale %>%
  filter(individual_sex=="F")%>%
  filter(!is.na(year))%>%
  filter(!is.na(effortSeason))%>%
  filter(!is.na(birthSznSD))%>%
  filter(!is.na(Jdate))%>%
  filter(!is.na(individual_code))

#create a table with the average survival organized by year and 'effortSeason'
#effortSeason = high vs. low reproductive investment
femYearData<-femYearData%>%
  group_by(year,effortSeason)%>%
  summarise(survival=mean(survival),n=n())

#plot Figure 2A
plot<-conditional_effects(survFemaleYear1,effects=c("year:effortSeason"))
femYear<-plot(plot)[[1]] +
  coord_cartesian(y=c(0.975,1))+
  geom_point(data=femYearData,aes(x=year,y=survival,color=effortSeason),inherit.aes=FALSE)+
  labs(x="Year", y="Survival Probability",title = "Female Survival")+
  scale_fill_manual(values=c("green4", "red3"),name = "Season", 
                    labels = c("Non-Birth", "Birth"))+
  scale_color_manual(values=c("green4", "red3"),name = "Season", 
                     labels = c("Non-Birth", "Birth"))+
  theme_classic()



#males - Model A Plot#########

#create dataframe that is complete for model
maleYearData<-survivalDataScale %>%
  filter(individual_sex=="M")%>%
  filter(!is.na(year))%>%
  filter(!is.na(effortSeason))%>%
  filter(!is.na(birthSznSD))%>%
  filter(!is.na(Jdate))%>%
  filter(!is.na(individual_code))

#create a table with the average survival organized by year and 'effortSeason'
maleYearData<-maleYearData%>%
  group_by(year,effortSeason)%>%
  summarise(survival=mean(survival),n=n())

#plot figure 2B
plot<-conditional_effects(survMaleYear1,effects=c("year:effortSeason"))
maleYear<-plot(plot)[[1]] +
  coord_cartesian(y=c(0.975,1))+
  geom_point(data=maleYearData,aes(x=year,y=survival,color=effortSeason),inherit.aes=FALSE)+
  labs(x="Year", y="Survival Probability",title = "Male Survival")+
  scale_fill_manual(values=c("purple4", "goldenrod"),name = "Season", 
                    labels = c("Non-Mating", "Mating"))+
  scale_color_manual(values=c("purple4", "goldenrod"),name = "Season", 
                     labels = c("Non-Mating", "Mating"))+
  theme_classic()
```

```{r Figure 2}
#combine male and female plots#############
#plot Figure 2
ggarrange(femYear, maleYear, common.legend=FALSE,legend="top",nrow=1,labels="AUTO")
```




# 4. Demographic effects on survival

We ran two models estimating the effects of demography on season-specific survival - one model for males and one for females.

## 4.1 Model Formulations and results

The variables included in the models are:

Response variable:

-   **survival:** A binomial variable denoting whether the individual survived (1) or died (0) during the 30 day period.

Fixed effects:

-   **JDate:** Number of days from January 1st until the month from which survival is measured. This is included as both first and second order terms using the 'poly()' function.

-   **effortSeason:** A 2-level categorical variable indicating if survival was measured during a month that majority overlapped with the individual's season of high reproductive investment ('yes') or not ('no').

-   **adultsTotal:** Population density, calculated as the total number of individuals above the age of one year alive on the island in a given month.

-   **age:** Individual age, included as both a 1st order and 2nd order polynomial term using the 'poly()' function. Note that the argument 'raw' in the poly() function is set to "TRUE" such that the age term is not standardized into orthogonal polynomials before the model is run. This is because the 'age' term, like all numeric fixed effects in the model, is already scaled and centered around 0 prior to analysis.

-   **propFemale:** Sex ratio, calculated as the proportion of the adult population (all individuals above the age of 4) that are female, in each given month.

Random effects:

-   **individual_code:** The unique identifier for the individual who's monthly survival is being measured.

-   **year:** The calendar year in which survival is being measured.

Below see the summary statistics and formulations for each model:

```{r demography model summaries}
#read in models
survFem3<-readRDS("survFem3.rds") # female demography model
survMale2B<-readRDS("survMale2B.rds") #male demography model


#the prior for all models was set as follows:
#weakly informative prior normally distributed w/ mean = 0, sd=1 for fixed effects & intercept
#random effect prior had standard deviation follow exp. distribution w/ a rate of 1
prior = c(prior(normal(0, 1), class = "b"),
  prior(normal(0, 1), class = "Intercept"),
 prior(exponential(1), class = "sd"))

#female demography model summary
survFem3$formula #model formula
print(summary(survFem3), digits = 4)

#male demography model summary
survMale2B$formula #model formula
print(summary(survMale2B), digits = 4)
```

## 4.2 Demographic effects graphs

Below we show the code to create Figure 3ABC in the manuscript. This figure illustrates the interaction effects of (A) age and reproductive investment, (B) sex ratio and reproductive investment, and (C) population density and reproductive investment, on female monthly survival probability.


```{r Figure 3 prep,message=FALSE,warning=FALSE,fig.show='hide'}
#female demographic graphs

#create dataframe that is complete for the model
femDemData<-survivalDataScale %>%
  filter(!is.na(adultsTotal))%>%
  filter(!is.na(effortSeason))%>%
  filter(!is.na(age))%>%
  filter(censorStatus=="dead")%>%
  filter(individual_sex=="F")%>%
  filter(!is.na(propFemale))%>%
  filter(!is.na(Jdate))%>%
  filter(!is.na(year))%>%
  filter(!is.na(individual_code))

#find average survival for binned age categories,for each season
femDemDataAge<-femDemData%>%
  mutate(age=round(age,digits=1))%>%
  group_by(age,effortSeason)%>%
  summarise(survival=mean(survival),n=n())

#plot age X season
plot<-conditional_effects(survFem3,effects=c("age:effortSeason"))
femAge<-plot(plot)[[1]] +
  coord_cartesian(x=c(-1.2,3.4),y=c(0.94,1))+
  geom_point(data=femDemDataAge,aes(x=age,y=survival,color=effortSeason),
             inherit.aes=FALSE)+
  labs(x="Age", y="Survival",title="Females")+
  scale_fill_manual(values=c("green4", "red3"),name = "Season", 
                    labels = c("Non-Birth", "Birth"))+
  scale_color_manual(values=c("green4", "red3"),name = "Season", 
                     labels = c("Non-Birth", "Birth"))+
  theme_classic()

#find average survival for binned sex ratio categories,for each season
femDemDataSR<-femDemData%>%
  mutate(propFemale=round(propFemale,digits=1))%>%
  group_by(propFemale,effortSeason)%>%
  summarise(survival=mean(survival),n=n())

#plot sex ratio X season
plot<-conditional_effects(survFem3,effects=c("propFemale:effortSeason"))
femSexRatio<-plot(plot)[[1]] +
  coord_cartesian(y=c(0.94,1))+
  geom_point(data=femDemDataSR,aes(x=propFemale,y=survival,color=effortSeason),
             inherit.aes=FALSE)+
  labs(x="Sex Ratio", y="Survival",title="Females")+
  scale_fill_manual(values=c("green4", "red3"),name = "Season", 
                    labels = c("Non-Birth", "Birth"))+
  scale_color_manual(values=c("green4", "red3"),name = "Season", 
                     labels = c("Non-Birth", "Birth"))+
  theme_classic()


#find average survival for binned population density categories,for each season
femDemDataPD<-femDemData%>%
  mutate(adultsTotal=round(adultsTotal,digits=1))%>%
  group_by(adultsTotal,effortSeason)%>%
  summarise(survival=mean(survival),n=n())

#plot population density X season
plot<-conditional_effects(survFem3,effects=c("adultsTotal:effortSeason"))
femPopDens<-plot(plot)[[1]] +
  coord_cartesian(y=c(0.94,1))+
  geom_point(data=femDemDataPD,aes(x=adultsTotal,y=survival,color=effortSeason),
             inherit.aes=FALSE)+
  labs(x="Population Density", y="Survival",title="Females")+
  scale_fill_manual(values=c("green4", "red3"),name = "Season", 
                    labels = c("Non-Birth", "Birth"))+
  scale_color_manual(values=c("green4", "red3"),name = "Season", 
                     labels = c("Non-Birth", "Birth"))+
  theme_classic()
```

Below we show the code to create Figure 3DEF in the manuscript. This figure illustrates the interaction effects of (D) age and reproductive investment, (E) sex ratio and reproductive investment, and (F) population density and reproductive investment, on male monthly survival probability.

```{r Figure 4 prep,message=FALSE,warning=FALSE,fig.show='hide'}
#male demographic graphs

#create dataframe that is complete for model
maleDemData<-survivalDataScale %>%
  filter(!is.na(adultsTotal))%>%
  filter(!is.na(effortSeason))%>%
  filter(!is.na(age))%>%
  filter(censorStatus=="dead")%>%
  filter(individual_sex=="M")%>%
  filter(!is.na(propFemale))%>%
  filter(!is.na(Jdate))%>%
  filter(!is.na(year))%>%
  filter(!is.na(individual_code))

#find average survival for binned age categories,for each season
maleDemDataAge<-maleDemData%>%
  mutate(age=round(age,digits=1))%>%
  group_by(age,effortSeason)%>%
  summarise(survival=mean(survival),n=n())

#plot age X season
plot<-conditional_effects(survMale2B,effects=c("age:effortSeason"))
maleAge<-plot(plot)[[1]] +
  coord_cartesian(x=c(-1.2,3.1),y=c(0.94,1))+
  geom_point(data=maleDemDataAge,aes(x=age,y=survival,color=effortSeason),
             inherit.aes=FALSE)+
  labs(x="Age", y="Survival",title="Males")+
  scale_fill_manual(values=c("purple4", "goldenrod"),name = "Season", 
                    labels = c("Non-Mating", "Mating"))+
  scale_color_manual(values=c("purple4", "goldenrod"),name = "Season", 
                     labels = c("Non-Mating", "Mating"))+
  theme_classic()


#find average survival for binned sex ratio categories,for each season
maleDemDataSR<-maleDemData%>%
  mutate(propFemale=round(propFemale,digits=1))%>%
  group_by(propFemale,effortSeason)%>%
  summarise(survival=mean(survival),n=n())

#plot sex ratio X season
plot<-conditional_effects(survMale2B,effects=c("propFemale:effortSeason"))
maleSexRatio<-plot(plot)[[1]] +
  coord_cartesian(y=c(0.94,1))+
  geom_point(data=maleDemDataSR,aes(x=propFemale,y=survival,color=effortSeason),
             inherit.aes=FALSE)+
  labs(x="Sex Ratio", y="Survival",title="Males")+
  scale_fill_manual(values=c("purple4", "goldenrod"),name = "Season",
                    labels = c("Non-Mating", "Mating"))+
  scale_color_manual(values=c("purple4", "goldenrod"),name = "Season",
                     labels = c("Non-Mating", "Mating"))+
  theme_classic()


#find average survival for binned population density categories,for each season
maleDemDataPD<-maleDemData%>%
  mutate(adultsTotal=round(adultsTotal,digits=1))%>%
  group_by(adultsTotal,effortSeason)%>%
  summarise(survival=mean(survival),n=n())

#plot population density X season
plot<-conditional_effects(survMale2B,effects=c("adultsTotal:effortSeason"))
malePopDens<-plot(plot)[[1]] +
  coord_cartesian(y=c(0.94,1))+
  geom_point(data=maleDemDataPD,aes(x=adultsTotal,y=survival,color=effortSeason),
             inherit.aes=FALSE)+
  labs(x="Population Density", y="Survival",title="Males")+  
  scale_fill_manual(values=c("purple4", "goldenrod"),name = "Season",
                    labels = c("Non-Mating", "Mating"))+
  scale_color_manual(values=c("purple4", "goldenrod"),name = "Season", 
                     labels = c("Non-Mating", "Mating"))+
  theme_classic()
```

```{r Figure 3}
#combine male and females to plot figure 3
ggarrange(femAge, femSexRatio,femPopDens,maleAge, maleSexRatio,malePopDens, common.legend=FALSE,legend="top",nrow=2,ncol = 3,labels="AUTO",font.label = c(size=20),hjust=-1.5)
```




# 5. Weather station data comparison

Temperature and rainfall data is freely available from all publicly funded weather stations in Puerto Rico through the National Oceanic and Atmospheric Administration (NOAA) website:
<https://www.ncei.noaa.gov/cdo-web/>

Before we run the climatic models, we need to check that the weather data used is an adequate representation of the weather conditions experienced on Cayo Santiago. The most complete weather data from the region of Puerto Rico nearest to Cayo Santiago is from the Rio Piedras weather station (41 km from Cayo Santiago). We chose this weather station over the closer weather station of Humacao (6 km from Cayo Santiago) since Humacao only has weather data covering about half of the study length. Here we compare the monthly mean temperatures and rainfall observed at the two stations during the years they do overlap, to ascertain if Rio Piedras station is an adequate approximation of the weather occurring on Cayo Santiago.

## 5.1 Weather station data cleaning

First we clean and combine the data from the two weather stations.

```{r weather station data cleaning}
#tidy up the Humacao weather station data
weatherHum<- weatherHum%>% 
 rename(prcpH=prcp,tmaxH=tmax)%>% #identify precipitation and max temps from Humacao
  select(date,prcpH,tmaxH)%>%
  mutate(date1=format(date, "%Y/%m"))

#combine weather data from Humacao and Rio Piedras
weatherComb<-left_join(weatherHum,select(weatherRP,date,prcp,tmax),by=c("date")) 

#group combined weather data by month
#summarise the mean rainfall and maximum temperature in each month
weatherComb<-weatherComb%>%
  group_by(date1)%>%
  summarise(prcpH=mean(prcpH),tmaxH=mean(tmaxH),prcp=mean(prcp),tmax=mean(tmax))
```

## 5.2 Compare Rio Piedras vs. Humacao rainfall

A simple linear regression shows that monthly average rainfall in Rio Piedras (prcp) is highly predictive of monthly average rainfall in Humacao (prcpH) (R\^2 = 0.41, p\<0.001).

```{r weather station rainfall compare}
#regression of Rio Peiedras precipitation predicting Humacao precipitation
summary(weather<-lm(prcpH~prcp,data=weatherComb))  
```

## 5.3 Compare Rio Piedras vs. Humacao temperature

A simple linear regression shows that monthly average maximum temperature (tmax) in Rio Piedras is highly predictive of monthly average maximum temperature in Humacao (tmaxH) (R\^2 = 0.60, p\<0.001)

```{r weather station temperature compare}
#regression of Rio Peiedras temperature predicting Humacao temperature
summary(weather1<-lm(tmaxH~tmax,data=weatherComb)) 
```

# 6. Climatic effects on survival

We ran two models estimating the effects of demography on season-specific survival - one model for males and one for females.

## 6.1 Model Formulations and results

The variables included in the models are:

Response variable:

-   **survival:** A binomial variable denoting whether the individual survived (1) or died (0) during the 30 day period.

Fixed effects:

-   **effortSeason:** A 2-level categorical variable indicating if survival was measured during a month that majority overlapped with the individual's season of high reproductive investment ('yes') or not ('no').

-   **adultsTotal:** Population density, calculated as the total number of individuals above the age of one year alive on the island in each given month

-   **age:** Individual age, included as both a 1st order and 2nd order polynomial term using the 'poly()' function. Note that the argument 'raw' in the poly() function is set to "TRUE" such that the age term is not standardized into orthogonal polynomials before the model is run. This is because the 'age' term, like all numeric fixed effects in the model, is already scaled and centered around 0 prior to analysis.

-   **propFemale:** Sex ratio, calculated as the proportion of the adult population (all individuals above the age of 4) that were female, in each given month.

-   **tmax:** Temperature, calculated as the average of all the daily maximum temperatures in centigrade recorded over the course of the month.

-   **rain:** Rainfall, calculated as the average of all the daily rainfall measures in millimeters recorded over the course of the month.

-   **JDate:** Number of days from January 1st until the month from which survival is measured. This is included as both first and second order terms using the 'poly()' function.

Random effects:

-   **individual_code:** The unique identifier for the individual who's monthly survival is being measured.

-   **year:** The calendar year in which survival is being measured.

Below see the summary statistics and formulations for each model:


```{r climate model summaries, warning=FALSE}
#read in models
survFemWeather1G<-readRDS("survFemWeather1G.rds") # female climate model
survMaleWeather1B<-readRDS("survMaleWeather1B.rds") # male climate model

#the prior for both models was set as follows:
#weakly informative prior normally distributed w/ mean = 0, sd=1 for intercept & fixed effects
prior = c(prior(normal(0, 1), class = "b"), 
  prior(normal(0, 1), class = "Intercept"), 
 prior(exponential(1), class = "sd")) 

#female climate model formula and summary
formula(survFemWeather1G)
print(summary(survFemWeather1G), digits = 4)

#male climate model formula and summary
formula(survMaleWeather1B)
print(summary(survMaleWeather1B), digits = 4)
```

## 6.2 Climatic effects graphs

Below we show the code to create Figure 4AB in the manuscript. This figure illustrates the interaction effects of (A) temperature and reproductive investment, and (B) rainfall and reproductive investment, on female monthly survival probability.

```{r Figure 4AB prep,message=FALSE,warning=FALSE,fig.show='hide'}
#female climate graphs

#check models estimates against raw data
#create dataframe that is complete for model
femWeatherData<-survivalDataScale  %>%
  filter(!is.na(age))%>%
  filter(!is.na(tmax))%>%
  filter(!is.na(rain))%>%
  filter(censorStatus=="dead")%>%
  filter(individual_sex=="F")%>%
  filter(!is.na(year))%>%
  filter(!is.na(individual_code))

#season X tmax
femWeatherDataTemp<-femWeatherData%>%
  mutate(tmax=round(tmax,digits=1))%>%
  group_by(tmax,effortSeason)%>%
  summarise(survival=mean(survival),n=n())
plot<-conditional_effects(survFemWeather1G,effects=c("tmax:effortSeason"))
femTemp<-plot(plot)[[1]] +
  coord_cartesian(x=c(-2.1,2),y=c(0.97,1))+
  geom_point(data=femWeatherDataTemp,aes(x=tmax,y=survival,color=effortSeason),inherit.aes=FALSE)+
  labs(x="Temperature", y="Survival", title="Females")+
  scale_fill_manual(values=c("green4", "red3"),name = "Season", labels = c("Non-Birth", "Birth"))+
  scale_color_manual(values=c("green4", "red3"),name = "Season", labels = c("Non-Birth", "Birth"))+
  theme_classic()

#season X rain
femWeatherDataRain<-femWeatherData%>%
  mutate(rain=round(rain,digits=1))%>%
  group_by(rain,effortSeason)%>%
  summarise(survival=mean(survival),n=n())
plot<-conditional_effects(survFemWeather1G,effects=c("rain:effortSeason"))
femRain<-plot(plot)[[1]] +
  coord_cartesian(x=c(-1.1,3),y=c(0.97,1))+
  geom_point(data=femWeatherDataRain,aes(x=rain,y=survival,color=effortSeason),inherit.aes=FALSE)+
  labs(x="Rainfall", y="Survival", title="Females")+
  scale_fill_manual(values=c("green4", "red3"),name = "Season", labels = c("Non-Birth", "Birth"))+
  scale_color_manual(values=c("green4", "red3"),name = "Season", labels = c("Non-Birth", "Birth"))+
  theme_classic()
```


Below we show the code to create Figure 4CD in the manuscript. This figure illustrates the interaction effects of (A) temperature and reproductive investment, and (B) rainfall and reproductive investment, on male monthly survival probability.


```{r Figure 4CD prep,message=FALSE,warning=FALSE,fig.show='hide'}
#male climate graphs


#check models estimates against raw data
#create dataframe that is complete for model
maleWeatherData<-survivalDataScale  %>%
  filter(!is.na(age))%>%
  filter(!is.na(tmax))%>%
  filter(!is.na(rain))%>%
  filter(censorStatus=="dead")%>%
  filter(individual_sex=="M")%>%
  filter(!is.na(year))%>%
  filter(!is.na(individual_code))

#season X tmax
maleWeatherDataTemp<-maleWeatherData%>%
  mutate(tmax=round(tmax,digits=1))%>%
  group_by(tmax,effortSeason)%>%
  summarise(survival=mean(survival),n=n())
plot<-conditional_effects(survMaleWeather1B,effects=c("tmax:effortSeason"))
maleTemp<-plot(plot)[[1]] +
  coord_cartesian(x=c(-2.1,2),y=c(0.97,1))+
  geom_point(data=maleWeatherDataTemp,aes(x=tmax,y=survival,color=effortSeason),inherit.aes=FALSE)+
  labs(x="Temperature", y="Survival", title="Males")+  
  scale_fill_manual(values=c("purple4", "goldenrod"),name = "Season", labels = c("Non-Mating", "Mating"))+
  scale_color_manual(values=c("purple4", "goldenrod"),name = "Season", labels = c("Non-Mating", "Mating"))+
  theme_classic()

#season X rain
maleWeatherDataRain<-maleWeatherData%>%
  mutate(rain=round(rain,digits=1))%>%
  group_by(rain,effortSeason)%>%
  summarise(survival=mean(survival),n=n())
plot<-conditional_effects(survMaleWeather1B,effects=c("rain:effortSeason"))
maleRain<-plot(plot)[[1]] +
  coord_cartesian(x=c(-1.1,3),y=c(0.97,1))+
  geom_point(data=maleWeatherDataRain,aes(x=rain,y=survival,color=effortSeason),inherit.aes=FALSE)+
  labs(x="Rainfall", y="Survival", title="Males")+ 
  scale_fill_manual(values=c("purple4", "goldenrod"),name = "Season", labels = c("Non-Mating", "Mating"))+
  scale_color_manual(values=c("purple4", "goldenrod"),name = "Season", labels = c("Non-Mating", "Mating"))+
  theme_classic()
```

```{r Figure 4}
#plot figure 4
ggarrange(femTemp, femRain, maleTemp, maleRain, common.legend=FALSE,legend="top",nrow=2,ncol = 2,labels="AUTO")
```
